Below-ground sink limitation alters growth and carbon balance of Eucalyptus seedlings

### 2015-09-30 manuscript/manuscript_pve.Rmd
edits to manuscript (including tables) from MT

#table data
library(pixiedust)
table1 <- read.csv("master_scripts/pve_table1.csv")
table2 <- read.csv("master_scripts/pve_table2.csv")

@@

\
Leaf dark respiration rates (R~d~) was measured on each seedling during the same dates as AC~i~ curves. Freshly detached leaves were collected at least 1 hour after sundown and placed inside a conifer chamber attached to the Licor 6400. Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of R~d~ are standardized rates at 25 &deg;C using a Q10 value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

@@

\
(1)
$$P_i = L\left(\frac{C_{day'i}~\sigma_s}{\epsilon_c}\right) - R$$
where L is total plant leaf area (m^2^), C~day,i~ is the predicted daily carbon assimilation (g C d^-1^), $\sigma$~s~ is a self shading parameter, $\epsilon$~c~ is a biomass conversion efficiency parameter and R is the mass based total respiration of all tissue components (g C d^-1^). Total respiration was calculated as

\
(2)
$$R = \Sigma(R_c~M_c)$$
where R~c~ is tissue respiration of fine roots, coarse roots or stems (g C d^-1^) and M~c~ is the standing biomass of each component (g). R~leaf~ is represented in the calculation of C~day~ (described below). The change in individual component biomass (M~c~), here solved on a daily time step, is given by

@@

\
We then utilized this model to test the hypothesis that the effects of belowground sink limitation on rates of leaf A where sufficient to accurately predict overall seedling biomass production after 120 days. Each model run utilized changes in A and leaf mass fraction (LMF), with published or local values of stem and root respiration rates, to generate total seedling mass and leaf area after 120 days.  Cumulative net leaf C gain for each treatment was equal to the sum of each value of C~day,i~ over 120 days and final seedling C was assumed to equal half of the final mass for both modeled and observed seedlings. First, a default model was optimized to produce a final LMF that correctly predicted both the final leaf mass and total biomass of the harvested free seedling controls (M~0~). This optimized LMF was then applied to model runs with treatment specific C~day~ to determine if changes in leaf A alone could accurately predict biomass (M~1~). Next, model sensitivity to different C allocation scenarios, including non-optimized treatment specific LMF and up regulation of non-leaf tissue respiration by 50 % of default values (M~2,3~, respectively), was used to improve predictions of initial model simulations from measured harvest biomass. For all cases, seedling biomass production was compared between model output and harvested seedlings with treatment specific mean daily C assimilation by first scaling values to the free seedling control.

##Data analysis
Differences in experimental parameters with soil volume were analysed by mixed-effects models in R with individual containers as random effects and soil volume treatment as a categorical fixed effect iwth seven levels. Tukey's post-hoc tests were performed in conjunction with ANOVA to determine which specific paired comparisons among soil volume treatments were different. A linear mixed effect model of A~max~ and leaf chemistry was performed using the 'nlme' package [@nlme] in R. Explained variance (R^2^) of mixed models were computed as in  [@nakagawa2013general]. Tests of allometric relationships between log-transformed biomass components were implemented using major axis regression in the 'smatr' package in R [@warton2012smatr]. Results were considered significant at P < 0.05.

@@

## Biomass partitioning under sink limitation
As biomass partitioning is likely controlled by the source and sink strength of all organs [@poorter2012biomass], it was important to determine which tissue components were most affected by the container size treatments. It was necessary to distinguish if growth was affected beyond ontogenetic constraints, by correcting for size, as biomass distribution is strongly size-dependent [@gould1966allometry; @lleonart2000removing]. In this study, there was no significant difference in root, leaf, or stem biomass partitioning with reduced soil volume compared to free seedlings, outside of ontogenetic drift. This is a surprising result as shifts in allocation have been noted specifically for nutrient limitation  [@mcconnaughay1999biomass, and references therein]. Surprisingly, fine root to leaf mass ratio was conserved across all treatments suggesting a functional partitioning response to optimize resource gain did not occur.

\
As partitioning to fine roots did not change this provides evidence against an optimal foraging strategy for seedlings in containers. This could be because lateral root development is affected by inanimate obstacles and avoiding growth towards container walls could improve the efficiency of resource allocation [@falik2005root]. Also, the sensitivity of roots to their own exudates near obstructions may be used to adjust growth before stressful conditions occur [@semchenko2008foraging]. Alternatively, physical restriction of root proliferation could have impacted root development and morphology prior to shifts in mass partitioning. Here, increases in root length where detected in several of the soil volume treatments. This is not surprising as plants in containers have been shown to have different root morphology to field grown plants as roots essentially compete among themselves for nutrients  [@nesmith1998effect]. The poor soil quality used in our experiment and root restriction, however, likely decreased the capacity of this morphological response to increase N uptake.

@@

\
**Table 2**. Responses of leaf level gas exchange parameters of *Eucalyptus tereticornis* seedlings to soil volume treatments. Each value reflects the mean (&pm; 1 standard error) for each treatment. Units for A~max~ and R~dark~ are $\mu$mol m^-2^ s^-1^ and g~s~ are mol m^-2^ s^-1^, each measured at 25 &deg;C. Values of A~max~, g~s~ and g~1~ represent overall means across measurement campaigns (n = 6). R~dark~, J~max~ and Vc~max~ values are means of two measurement campaigns at beginning and end of gas exchange measurements. Different letters represent significant differences between treatments. The container effect P value represents the overall difference between seedlings with soil volume restriction and the free seedling treatment.

@@

dust(table1)+
  sprinkle_colnames("Volume (L)", 
                        "Seedling~mass (g)", 
                        "SLA TNC-free (m^2^ kg^-1^)",
                        "Leaf Starch (%)", 
                        "Leaf Sugars (%)", 
                        "Leaf Nitrogen (%)",
                        "Root Nitrogen (%)",
                        "SRL (cm m^-1^)",
                        paste0("Leaf ", greeks['delta'],"^13^C"," (\u2030)")) +
  sprinkle(rows= 1, part="head",bold=TRUE)+
  sprinkle_print_method("markdown")
```

\
**Table 2.** Responses of leaf level gas exchange parameters of *Eucalyptus tereticornis* seedlings to soil volume treatments. Each value reflects the mean(standard error) for each treatment. Units for A~max~ and R~dark~ are $\mu$mol m^-2^ s^-1^ and g~s~ are mol m^-2^ s^-1^, each at at 25 &deg;C. Values of A~max~, g~s~ and g~1~ represent overall means across measurement campaigns (n=6). R~dark~, J~max~ and Vc~max~ values are means of two measurement campaigns at beginning and end of gas exchange measurements. Different letters represent significant differences between treatments. The container effect P value represents the overall difference between seedlings with soil volume restriction and the control seedlings.

```{r table2, echo=FALSE, warning=FALSE}
dust(table2)+
  sprinkle_colnames("Volume (L)", 
                        "A~max~", 
                        "R~dark~",
                        "J~max~",
                        "Vc~max~",
                        "g~s~",
                        "g~1~") +
  sprinkle(rows= 1, part="head",bold=TRUE)+
  sprinkle_print_method("markdown")
```

@@

**Figure S1**.

>>>>

#table data
library(pixiedust)
table1 <- read.csv("master_scripts/pve_table1.csv")
table2 <- read.csv("master_scripts/pve_table2.csv")
modeltable <- read.csv("master_scripts/modeltable.csv")

@@

\
Leaf dark respiration rates (R) was measured on each seedling during the same dates as AC~i~ curves. Freshly detached leaves were collected at least 1 hour after sundown and placed inside a conifer chamber attached to the Licor 6400. Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of R are standardized rates at 25 &deg;C using a Q10 value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

@@

\
(1)
$$P_i = L\left(\frac{C_{day'i}~\sigma_s}{\epsilon_c}\right) - R$$
where L is total plant leaf area (m^2^), C~day,i~ is the predicted daily carbon assimilation (g C g mass^-1^ d^-1^), $\sigma$~s~ is a self shading parameter, $\epsilon$~c~ is a biomass conversion efficiency parameter and R is the mass based total respiration of all tissue components (g C d^-1^). Total respiration was calculated as

\
(2)
$$R = \Sigma(R_c~M_c)$$
where R~c~ is publsihed or local tissue respiration rates of fine roots, coarse roots or stems (Table S1, g C g mass ^-1^ d^-1^) and M~c~ is the standing biomass of each component (g). R~leaf~ is represented in the calculation of C~day~ (described below). The change in individual component biomass (M~c~), here solved on a daily time step, is given by

@@

\
We then utilized this model to test the hypothesis that the effects of belowground sink limitation on rates of leaf A where sufficient to accurately predict overall seedling biomass production after 120 days. Each model run utilized changes in A and leaf mass fraction (LMF), with values of stem and root respiration rates, to generate total seedling mass and leaf area after 120 days.  Cumulative net leaf C gain for each treatment was equal to the sum of each value of C~day,i~ over 120 days and final seedling C was assumed to equal half of the final mass for both modeled and observed seedlings. First, a default model was optimized to produce a final LMF that correctly predicted both the final leaf mass and total biomass of the harvested free seedling controls (M~0~). This optimized LMF was then applied to model runs with treatment specific C~day~ to determine if changes in leaf A alone could accurately predict biomass (M~1~). Next, model sensitivity to different C allocation scenarios, including non-optimized treatment specific LMF and up regulation of non-leaf tissue respiration by 50 % of default values (M~2,3~, respectively), was used to improve predictions of initial model simulations from measured harvest biomass. For all cases, seedling biomass production was compared between model output and harvested seedlings with treatment specific mean daily C assimilation by first scaling values to the free seedling control.

##Data analysis
Differences in experimental parameters with soil volume were analysed by mixed-effects models in R with individual containers as random effects and soil volume treatment as a categorical fixed effect iwth seven levels. Tukey's post-hoc tests were performed in conjunction with ANOVA to determine which specific paired comparisons among soil volume treatments were different. A linear mixed effect model of A~max~ and leaf chemistry was performed using the 'nlme' package [@nlme] in R. Explained variance (R^2^) of mixed models were computed as in  [@nakagawa2013general]. Tests of allometric relationships between log-transformed biomass components were implemented using standardized major axis regression in the 'smatr' package in R [@warton2012smatr]. Results were considered significant at P < 0.05.

@@

## Biomass partitioning under sink limitation
As biomass partitioning is likely controlled by the source and sink strength of all organs [@poorter2012biomass], it was important to determine which tissue components were most affected by the container size treatments. It was necessary to distinguish if growth was affected beyond ontogenetic constraints, by correcting for size, as biomass distribution is strongly size-dependent [@gould1966allometry; @lleonart2000removing]. In this study, there was no significant difference in root, leaf, or stem biomass partitioning with reduced soil volume compared to free seedlings, outside of ontogenetic drift. This is a surprising result as shifts in allocation have been noted specifically for nutrient limitation  [@mcconnaughay1999biomass, and references therein]. Surprisingly, a constant ratio of fine root mass to leaf mass was observed across all treatments suggesting a functional partitioning response to optimize resource gain did not occur.

\
As partitioning to fine roots did not change this provides evidence against an optimal foraging strategy for seedlings in containers. This could be because lateral root development is affected by inanimate obstacles and avoiding growth towards container walls could improve the efficiency of resource allocation [@falik2005root]. Also, the sensitivity of roots to their own exudates near obstructions may be used to adjust growth before stressful conditions occur [@semchenko2008foraging]. Alternatively, physical restriction of root proliferation could have impacted root development and morphology prior to shifts in mass partitioning. Here, increases in root length where detected in several of the soil volume treatments. This is not surprising as plants in containers have been shown to have different root morphology to field grown plants [@nesmith1998effect]. The poor soil quality used in our experiment and root restriction, however, likely decreased the capacity of this morphological response to increase N uptake.

@@

\
**Table 2.** Responses of leaf level gas exchange parameters of *Eucalyptus tereticornis* seedlings to soil volume treatments. Each value reflects the mean(standard error) for each treatment. A~max~, R~dark~ and g~s~ are each measured at 25 &deg;C. Values of A~max~, g~s~ and g~1~ represent overall means across measurement campaigns (n=6). R~dark~, J~max~ and Vc~max~ values are means of two measurement campaigns at beginning and end of gas exchange measurements. Different letters represent significant differences between treatments. The container effect P value represents the overall difference between seedlings with soil volume restriction and the control seedlings.

@@

dust(table1)+
  sprinkle_colnames("Volume (L)", 
                        "Seedling mass (g)", 
                        "SLA~TNC-free~ (m^2^ kg^-1^)",
                        "Leaf Starch (%)", 
                        "Leaf Sugars (%)", 
                        "Leaf Nitrogen (%)",
                        "Root Nitrogen (%)",
                        "SRL (cm m^-1^)",
                        paste0("Leaf ", greeks['delta'],"^13^C"," (\u2030)")) +
  sprinkle(rows= 1, part="head",bold=TRUE)+
  sprinkle_print_method("markdown")
```

\
**Table 2.** Responses of leaf level gas exchange parameters of *Eucalyptus tereticornis* seedlings to soil volume treatments. Each value reflects the mean(standard error) for each treatment. A~max~, R~dark~ and g~s~ are each measured at 25 &deg;C. Values of A~max~, g~s~ and g~1~ represent overall means across measurement campaigns (n=6). R~dark~, J~max~ and Vc~max~ values are means of two measurement campaigns at beginning and end of gas exchange measurements. Different letters represent significant differences between treatments. The container effect P value represents the overall difference between seedlings with soil volume restriction and the control seedlings.

```{r table2, echo=FALSE, warning=FALSE}
dust(table2)+
  sprinkle_colnames("Volume (L)", 
                        "A~max~ ($\\mu$mol m^-2^ s^-1^)", 
                        "R~dark~ ($\\mu$mol m^-2^ s^-1^)",
                        "J~max~",
                        "Vc~max~",
                        "g~s~ (mol m^-2^ s^-1^)",
                        "g~1~") +
  sprinkle(rows= 1, part="head",bold=TRUE)+
  sprinkle_print_method("markdown")
```

@@

**Figure S1**. Seedling growth model default parameters.
```{r tabels1, echo=FALSE, warning=FALSE}
dust(modeltable)+
  sprinkle(rows= 1, part="head",bold=TRUE)+
  sprinkle_print_method("markdown")
```
